Psychological Needs During Intergroup Contact

### 2022-01-17 Methods-and-Results.tex
re-read and tbl/fig references

After receiving ethical approval from the University of Groningen, we
recruited 23 migrants using the local paid participant pool and
specifically targeted non-Dutch migrants to participate in our study.
Participants reported on their interactions for at least 30 days with
two daily measures (capturing the morning and afternoon). With this
design, we aimed at getting 50-60 measurements per participant
(\textit{M} = 53.26, \textit{SD} = 16.72, \textit{total N} = 1225). This
is a common number of measurements found in experience sampling studies
and offers sufficient power to model processes within and between
participants \citep[e.g., for a systematic review see][]{AanhetRot2012}.
Participants were compensated for their participation with up to 34
Euros -- each two Euros for pre- and post-questionnaire as well as 50
Eurocents for every experience sampling measurement occasion. The sample
consisted of relatively young, educated, and western migrants from the
global north (\(M_{age}\) = 24.35, \(SD_{age}\) = 4.73, 19 women, 15
students). The sample accurately describes one of the largest groups of
migrants in the region \citep[][]{GemeenteGroningen2015}.

@@

To test the prerequisite effect of intergroup contact, every experience
sampling measurement started with the question
``\textit{Did you meet a Dutch person this morning [/afternoon]? (In person interaction for at least 10 minutes)}''.
Our participants recorded between 2--51 interactions with Dutch outgroup
members (3.23--91.07\% of individual daily diary measurements; 31.59\%
of all 1225 daily diary
responses)\footnote{Two participants only recorded two daily diary measurements each and non of these included outgroup contacts. These participants are removed from any analyses including outgroup contacts.}.

@@

We tested the most general contact hypothesis in two steps. First, we
assessed whether more intergroup interactions were related to to more
positive outgroup attitudes. Second, we tested whether a potential
positive effect on outgroup attitudes depended on the interaction
quality (jointly with the number of interactions). We find that neither
the number of interactions nor the number of daily diary responses with
an interaction were significantly related with the average outgroup
attitudes (all \textbar{} \textit{r} \textbar{} \textless{} 0.03, all
\textit{p} \textgreater{} 0.887). However, we did find a medium sized
correlation between the participants' Average Interaction Quality and
their Average Outgroup Attitudes (\textit{r} = \hl{X.XX}, \textit{p} =
\hl{X.XXX}. And when considering the number of interactions and average
interaction quality jointly in a linear regression, we additionally find
a statistically significant interaction term (\textit{b} = -0.05,
\textit{t}(17) = -2.95, \textit{p} = 0.009, \(\eta_p^2\) = 0.34), where
in our sample with an increasing number of interactions the positive
effect of average interaction quality becomes weaker. We additionally
used a multilevel regression to check whether having an interaction with
an outgroup member had a situational (i.e., contemporaneous) effect
within the participants. We find that having an outgroup interaction is
indeed associated with significantly more positive outgroup attitudes
within the participants (\textit{b} = 2.48, \textit{t}(1200) = 4.36,
\textit{p} \textless{} .001, \textit{95\%CI}{[}1.36, 3.59{]}), even
after controlling for having an interaction with a non-Dutch (which did
not relate to outgroup attitudes independently; For full results see
Online Supplementary Materials
B)\footnote{Interestingly, adding random slopes to this model did not explain additional variance. This is unusual and might indicate the the effect is very consistent across participants. However, the small number of participants, or other measurement issues provide an alternative explanation}.
Thus, in our first data we find mixed results, where outgroup contacts
show a positive effect on outgroup attitudes in the moment but these
results do not translate into average differences between participants
(eventough the average number of interactions does seem to interact with
average interaction quality to predict outgroup attitudes). However, it
should be noted that the between-participant analyses in the aggregated
form are based on a small sample size.

@@

The main proposal of our article is that the success of an outgroup
interaction might be explained by whether or not the interaction
fulfilled the person's core situational need. This should, in turn, be
due to a higher perceived interaction quality. We sequentially test
whether the fulfillment of the core need during an interaction is (1)
related to more positive outgroup attitudes, (2) higher perceived
interaction quality, and (3) whether the variance explained by the core
need is assumed by the perceived interaction quality if considered
jointly.

@@

To build further confidence in our results, we assessed two additional
models that might offer alternative explanations. First, to ensure that
the effect of core need fulfillment is specific to an actual
interaction, we compared the effect to core need fulfillement in
situations without an interaction. For this, we analyzed the generalized
situational core need fulfillment (either during an interaction or about
the daytime in general) and tested whether the effect differed during
daily diary measurements with and without outgroup contacts. We found no
main effect of core need fulfillment (random slopes model; \textit{b} =
-0.10, \textit{t}(1199) = -2.22, \textit{p} = 0.026,
\textit{95\%CI}{[}-0.18, -0.01{]}) but a significant interaction effect
of core need fulfillment and outgroup contact (\textit{b} = 0.13,
\textit{t}(1199) = 4.31, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.07, 0.18{]}). Together with a significant main
effect of having an outgroup contact, this indicates that it is not key
need fulfillment in general --- but only key need fulfillment during an
outgroup contact that predicts more positive outgroup attitudes.

In a final step, we controlled for other fundamental psychological needs
during the interaction. We focus on the three commonly considered self
determination needs: competence, autonomy, and relatedness. We find that
the core need adds significantly above a model with only the self
determination theory needs (random slopes models; \(\chi^2\)(6,
\textit{N} = 21) = 19.60, \textit{p} = 0.003). We find that next to
relatedness (\textit{b} = 0.10, \textit{t}(361) = 4.04, \textit{p}
\textless{} .001, \textit{95\%CI}{[}0.05, 0.14{]}), the core need
explains the most variance in outgroup attitudes after an outgroup
contact (\textit{b} = 0.09, \textit{t}(361) = 2.23, \textit{p} = 0.026,
\textit{95\%CI}{[}0.01, 0.18{]}). When compared to the model with only
the SDT needs, the core need fulfillment flexibly takes on some of the
explained variance of all three fundamental needs (competence and
autonomy needs turning non-significant; all \textit{b} \textless{} 0.05,
all \textit{p} \textgreater{} 0.544). For full results see Online
Supplementary Information B.

@@

We tested the most general contact hypothesis in the same manner as in
Study 1. We found a significant correlation with average outgroup
attitudes for both the total number of outgroup interactions (\textit{r}
= 0.22, \textit{p} = 0.019) and the number of measurement beeps with an
interaction (\textit{r} = 0.28, \textit{p} = 0.003). However, we found
no significant correlation between the participants' Average Interaction
Quality and their Average Outgroup Attitudes (\textit{r} = 0.04,
\textit{p} = 0.679). Thus, when considering the number of interactions
and average interaction quality jointly in a linear regression, we find
a main effect of intergroup contacts to predict outgroup attitudes
(\textit{b} = 0.81, \textit{t}(109) = 3.29, \textit{p} = 0.001,
\(\eta_p^2\) = 0.08) but no significant effect of average perceived
contact quality and no significant interaction term. We additionally
used a multilevel regression to check whether having an interaction with
an outgroup member had a situational (i.e., contemporaneous) effect
within the participants. We find that having an outgroup interaction is
indeed associated with significantly more positive outgroup attitudes
within the participants (random slopes model; \textit{b} = 2.99,
\textit{t}(4850) = 3.80, \textit{p} \textless{} .001,
\textit{95\%CI}{[}1.45, 4.54{]}), even after controlling for having an
interaction with a non-Dutch person (which did not relate to outgroup
attitudes independently; For full results see Online Supplementary
Materials B). Thus, in our second dataset we find mixed results,
outgroup contacts show a positive effect on outgroup attitudes in the
moment and on average between participants. However, we find that this
effect does not depend on the average perceived interaction quality.
This unexpected results, might be due to the aggregation process and the
following analyses will focus in on the role of the perceived
interaction quality during an outgroup interaction.

@@

We again sequentially tested whether the fulfillment of the core need
during an interaction was (1) related to more positive outgroup
attitudes, (2) higher perceived interaction quality, and (3) whether the
variance explained by the core need is assumed by the perceived
interaction quality if considered jointly. We find that in the
multilevel models, the fulfillment of core situational needs during
outgroup contacts was associated with more positive outgroup attitudes
(random slopes model; \textit{b} = 0.13, \textit{t}(826) = 4.18,
\textit{p} \textless{} .001, \textit{95\%CI}{[}0.07, 0.19{]}) and also
predicted higher perceived interaction quality (random slopes model;
\textit{b} = 0.40, \textit{t}(826) = 6.46, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.28, 0.52{]}). Additionally, if we consider the
influences of core need fulfillment and interaction quality on outgroup
attitudes jointly, we find that virtually all variance is explained by
perceived interaction quality (random slopes mode; \textit{b} = 0.17,
\textit{t}(825) = 5.56, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.11, 0.23{]}) and only little unique variance is
still explained by core need fulfillment (\textit{b} = 0.03,
\textit{t}(825) = 1.47, \textit{p} = 0.141, \textit{95\%CI}{[}-0.01,
0.07{]}). These results are consistent with the results in study one and
we thus find support for our hypotheses and can conclude that in this
data set the fulfillment of core situational needs had a significant
influence on outgroup attitudes and that this effect is likely explained
by its effect through perceived interaction quality.

@@

We again checked for alternative models. First, when considering
generalized situational core need fulfillment together with whether an
intergroup contact took place, we find that there is only a minuscule
main effect of core need fulfillment (random slopes model; \textit{b} =
-0.03, \textit{t}(4849) = -1.34, \textit{p} = 0.181,
\textit{95\%CI}{[}-0.08, 0.02{]}) but a stronger interaction effect of
core need fulfillment and outgroup contact (\textit{b} = 0.06,
\textit{t}(4849) = 3.03, \textit{p} = 0.002, \textit{95\%CI}{[}0.02,
0.10{]}). Together with a significant main effect of having an outgroup
contact (\textit{b} = 2.88, \textit{t}(4849) = 3.71, \textit{p}
\textless{} .001, \textit{95\%CI}{[}1.36, 4.40{]}), this indicates that
it is not key need fulfillment in general --- but key need fulfillment
during an outgroup contact that predicts more positive outgroup
attitudes. This finding is consistent with the results of the previous
study, albeit with slightly weaker effect (likely because of the large
number of measurements that did not include an outgroup interaction).

In a final step we again checked whether during the interaction the core
situational need remains a meaningful predictor even when taking other
fundamental psychological needs into account. We find that the core need
adds significantly above a model with only the self determination theory
needs (random slopes models; \(\chi^2\)(6, \textit{N} = 108) = 22.90,
\textit{p} \textless{} .001). We find that the core need explains the
most variance in outgroup attitudes after an outgroup contact
(\textit{b} = 0.08, \textit{t}(823) = 2.95, \textit{p} = 0.003,
\textit{95\%CI}{[}0.03, 0.13{]}). When compared to the model with only
the SDT needs, the core need fulfillment flexibly takes on some of the
explained variance of all of the three fundamental needs. However,
different from the first study, relatedness (\textit{b} = 0.07,
\textit{t}(823) = 4.26, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.04, 0.10{]}) and autonomy (\textit{b} = 0.02,
\textit{t}(823) = 0.93, \textit{p} = 0.353, \textit{95\%CI}{[}-0.02,
0.08{]}) also predicted positive outgroup attitudes in this larger
sample. For full results see Online Supplementary Information B.

@@

As in Studies 1 and 2, we first conducted a general contact hypothesis
test. We found no correlation between intergroup contact and average
outgroup attitudes --- for neither the total number of outgroup
interactions nor the number of measurement beeps with an interaction
(\textit{r} \textless{} 0.13, \textit{p} \textgreater{} 0.291). However,
we did find a significant correlation between the participants' Average
Interaction Quality and their Average Outgroup Attitudes (\textit{r} =
0.25, \textit{p} = 0.036). When considering the number of interactions
and average interaction quality jointly in a linear regression, we only
found a main effect of average perceived interaction quality on outgroup
attitudes (\textit{b} = 0.36, \textit{t}(67) = 2.22, \textit{p} = 0.030,
\(\eta_p^2\) = 0.07) but no significant effect of the number of contacts
nor an interaction effect. This result mirrors that of study one but is
inconsistent with the second study. Additionally in a multilevel
regression to check for contemporaneous within person effects, we found
that having an outgroup interaction was indeed associated with
significantly more positive outgroup attitudes within the participants
(random slopes model; \textit{b} = 5.63, \textit{t}(3834) = 6.61,
\textit{p} \textless{} .001, \textit{95\%CI}{[}3.97, 7.29{]}), even
after controlling for having an interaction with a non-Dutch (which did
not relate to outgroup attitudes independently; for full results see
Online Supplementary Materials B). Thus, in our third data set we again
found mixed results. Outgroup contacts were on average not significantly
related to outgroup attitudes on a between participant level and this
effect was also not suppressed by an average interaction quality (i.e.,
no interaction effect). However, while the aggregate data was partly
inconsistent with the first two studies, the within person
contemporaneous effect of intergroup contact was consistent across all
three studies.

@@

We sequentially tested whether the fulfillment of Allport's Contact
Conditions was (1) related to more positive outgroup attitudes, (2)
higher perceived interaction quality, and (3) whether the variance
explained by Allport's Conditions is assumed by the perceived
interaction quality if considered jointly. We find that in the
multilevel models, the fulfillment of Allport's Conditions during
outgroup contacts was associated with more positive outgroup attitudes
(random slopes model; \textit{b} = 0.22, \textit{t}(1601) = 5.86,
\textit{p} \textless{} .001, \textit{95\%CI}{[}0.15, 0.29{]}) and also
predicted higher perceived interaction quality (random slopes model;
\textit{b} = 0.65, \textit{t}(1605) = 11.82, \textit{p} \textless{}
.001, \textit{95\%CI}{[}0.54, 0.76{]}). Moreover, when we consider the
influences of Allport's Conditions and interaction quality on outgroup
attitudes jointly, we find that perceived interaction quality is a
substantially stronger predictor (random slopes mode; \textit{b} = 0.17,
\textit{t}(1600) = 6.25, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.12, 0.23{]}) and the unique variance explained by
Allport's Conditions was less than half of its original effect size
(\textit{b} = 0.09, \textit{t}(1600) = 2.89, \textit{p} = 0.004,
\textit{95\%CI}{[}0.03, 0.15{]}). These results indicate that in this
dataset the the fulfillment of Allport's conditions had a significant
influence on outgroup attitudes and this effect is likely in parts
explained by its effect through perceived interaction quality.

@@

We again sequentially tested whether the fulfillment of the core need
during an interaction was (1) related to more positive outgroup
attitudes, (2) higher perceived interaction quality, and (3) whether the
variance explained by the core need is assumed by the perceived
interaction quality if considered jointly. We find that in the
multilevel models, the fulfillment of core need fulfillment during
outgroup contacts was associated with more positive outgroup attitudes
(random slopes model; \textit{b} = 0.19, \textit{t}(1601) = 5.29,
\textit{p} \textless{} .001, \textit{95\%CI}{[}0.12, 0.27{]}) and also
predicted higher perceived interaction quality (random slopes model;
\textit{b} = 0.43, \textit{t}(1605) = 8.35, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.33, 0.54{]}). Additionally, once we consider the
influences of core need fulfillment and interaction quality on outgroup
attitudes jointly, we find that perceived interaction quality is a
substantially stronger predictor (random slopes mode; \textit{b} = 0.17,
\textit{t}(1600) = 6.86, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.12, 0.22{]}) and the unique variance explained by
core need fulfillment was roughly half of its original effect size
(\textit{b} = 0.11, \textit{t}(1600) = 3.50, \textit{p} \textless{}
.001, \textit{95\%CI}{[}0.05, 0.17{]}). These results indicate that in
this dataset the the fulfillment of core situational needs had a
significant influence on outgroup attitudes and this effect is likely in
parts explained by its effect through perceived interaction quality.

@@

To test wether Allport's conditions or the core need fulfillment are
better at predicting outgroup attitudes, we first assess relative model
performance indices (i.e., Akaike information criterion, and Bayesian
information criterion), and then consider the two predictors in a joint
model to see whether the two approaches predict the same variance in
outgroup attitudes. When comparing the model selection indices we find
that the fulfillment of the situational core need, indeed performs
slightly better than the model using Allport's conditions
(\(AIC_{CoreNeed}\) 12632.02 \textless{} 12651.59 \(AIC_{Allport}\), and
\(BIC_{CoreNeed}\) 12664.55 \textless{} 12684.12 \(BIC_{Allport}\)).
Additionally, when considering the predictors jointly, we find that both
significantly predict outgroup attitudes with similar sized regression
parameters (random slopes model; Allport's Conditions: \textit{b} =
0.16, \textit{t}(1600) = 4.92, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.09, 0.24{]}, Core Need: \textit{b} = 0.14,
\textit{t}(1600) = 3.85, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.08, 0.17{]}). This indicates that, although both
Allport's conditions and the core need fulfillment seem to (in part)
work through perceived interaction quality, they explain different
aspects of the variance in outgroup attitudes and do not constitute one
another.

>>>>

After receiving ethical approval from the University of Groningen, we
recruited 23 migrants using the local paid participant pool and
specifically targeted non-Dutch migrants to participate in our study.
Participants reported on their interactions for at least 30 days with
two daily measures (capturing the morning and afternoon). With this
design, we aimed at getting 50-60 measurements per participant
(\textit{M} = 53.26, \textit{SD} = 16.72, \textit{total N} = 1225). This
is a common number of measurements found in experience sampling studies
and offers sufficient power to model processes within and between
participants \citep[e.g., for a systematic review see][]{AanhetRot2012}.
Participants were compensated for their participation with up to 34
Euros -- each two Euros for pre- and post-questionnaire as well as 50
Eurocents for every experience sampling measurement occasion. The sample
consisted of relatively young, educated, and western migrants from the
global north (\(M_{age}\) = 24.35, \(SD_{age}\) = 4.73, 19 women, 15
students). The sample accurately describes the largest groups of
migrants in the region \citep[][]{GemeenteGroningen2015}.

@@

To test the prerequisite effect of intergroup contact, every experience
sampling measurement started with the question
``\textit{Did you meet a Dutch person this morning [/afternoon]? (In person interaction for at least 10 minutes)}''.
Our participants recorded between 2--51 interactions with Dutch outgroup
members (3.23--91.07\% of individual daily diary measurements; 31.59\%
of all 1225 daily diary
responses)\footnote{Two participants only recorded two daily diary measurements each and non of these included outgroup contacts. These participants are removed from any analyses that focus on outgroup contacts.}.

@@

We tested the most general contact hypothesis in two steps. First, we
assessed whether more intergroup interactions were related to to more
positive outgroup attitudes. Second, we tested whether a potential
positive effect on outgroup attitudes depended on the interaction
quality (jointly with the number of interactions). We find that neither
the number of interactions nor the number of daily diary responses with
an interaction were significantly related with the average outgroup
attitudes (all \textbar{} \textit{r} \textbar{} \textless{} 0.03, all
\textit{p} \textgreater{} 0.887). However, we did find a medium sized
correlation between the participants' Average Interaction Quality and
their Average Outgroup Attitudes (\textit{r} = \hl{X.XX}, \textit{p} =
\hl{X.XXX}. And when considering the number of interactions and average
interaction quality jointly in a linear regression, we additionally find
a statistically significant interaction term (\textit{b} = -0.05,
\textit{t}(17) = -2.95, \textit{p} = 0.009, \(\eta_p^2\) = 0.34), where
in our sample with an increasing number of interactions the positive
effect of average interaction quality becomes weaker (also see Table
\ref{tab:intergroupGeneralTblWide}). We additionally used a multilevel
regression to check whether having an interaction with an outgroup
member had a situational (i.e., contemporaneous) effect within the
participants. We find that having an outgroup interaction is indeed
associated with significantly more positive outgroup attitudes within
the participants (\textit{b} = 2.48, \textit{t}(1200) = 4.36, \textit{p}
\textless{} .001, \textit{95\%CI}{[}1.36, 3.59{]}), even after
controlling for having an interaction with a non-Dutch (which did not
relate to outgroup attitudes independently; For full results see Table
\ref{tab:intergroupGeneralTblWide} and Online Supplementary Materials
B)\footnote{Interestingly, adding random slopes to this model did not explain additional variance. This is unusual and might indicate the the effect is very consistent across participants. However, the small number of participants, or other measurement issues provide an alternative explanation}.
Thus, in our first data we find mixed results, where outgroup contacts
show a positive effect on outgroup attitudes in the moment but these
results do not translate into average differences between participants
(eventough the average number of interactions does seem to interact with
average interaction quality to predict outgroup attitudes). However, it
should be noted that the between-participant analyses in the aggregated
form are based on a small sample size.

@@

The main proposal of our article is that the success of an outgroup
interaction might be explained by whether or not the interaction
fulfilled the person's core situational need. This should, in turn, be
due to a higher perceived interaction quality. We sequentially test
whether the fulfillment of the core need during an interaction is (1)
related to more positive outgroup attitudes, (2) higher perceived
interaction quality, and (3) whether the variance explained by the core
need is assumed by the perceived interaction quality if considered
jointly (for full regression results see Table
\ref{tab:intergroupNeedsTblWide}).

@@

To build further confidence in our results, we assessed two additional
models that might offer alternative explanations. First, to ensure that
the effect of core need fulfillment is specific to an actual
interaction, we compared the effect to core need fulfillement in
situations without an interaction. For this, we analyzed the generalized
situational core need fulfillment (either during an interaction or about
the daytime in general) and tested whether the effect differed during
daily diary measurements with and without outgroup contacts. We found no
main effect of core need fulfillment (random slopes model; \textit{b} =
-0.10, \textit{t}(1199) = -2.22, \textit{p} = 0.026,
\textit{95\%CI}{[}-0.18, -0.01{]}) but a significant interaction effect
of core need fulfillment and outgroup contact (\textit{b} = 0.13,
\textit{t}(1199) = 4.31, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.07, 0.18{]}; also see Table
\ref{tab:robustnessTblWide}). Together with a significant main effect of
having an outgroup contact, this indicates that it is not key need
fulfillment in general --- but only key need fulfillment during an
outgroup contact that predicts more positive outgroup attitudes.

In a final step, we controlled for other fundamental psychological needs
during the interaction. We focus on the three commonly considered self
determination needs: competence, autonomy, and relatedness. We find that
the core need adds significantly above a model with only the self
determination theory needs (random slopes models; \(\chi^2\)(6,
\textit{N} = 21) = 19.60, \textit{p} = 0.003). We find that next to
relatedness (\textit{b} = 0.10, \textit{t}(361) = 4.04, \textit{p}
\textless{} .001, \textit{95\%CI}{[}0.05, 0.14{]}), the core need
explains the most variance in outgroup attitudes after an outgroup
contact (\textit{b} = 0.09, \textit{t}(361) = 2.23, \textit{p} = 0.026,
\textit{95\%CI}{[}0.01, 0.18{]}). When compared to the model with only
the SDT needs, the core need fulfillment flexibly takes on some of the
explained variance of all three fundamental needs (competence and
autonomy needs turning non-significant; all \textit{b} \textless{} 0.05,
all \textit{p} \textgreater{} 0.544). For full results see see Table
\ref{tab:robustnessTblWide} and Online Supplementary Information B.

@@

We tested the most general contact hypothesis in the same manner as in
Study 1. We found a significant correlation with average outgroup
attitudes for both the total number of outgroup interactions (\textit{r}
= 0.22, \textit{p} = 0.019) and the number of measurement beeps with an
interaction (\textit{r} = 0.28, \textit{p} = 0.003). However, we found
no significant correlation between the participants' Average Interaction
Quality and their Average Outgroup Attitudes (\textit{r} = 0.04,
\textit{p} = 0.679). Thus, when considering the number of interactions
and average interaction quality jointly in a linear regression, we find
a main effect of intergroup contacts to predict outgroup attitudes
(\textit{b} = 0.81, \textit{t}(109) = 3.29, \textit{p} = 0.001,
\(\eta_p^2\) = 0.08) but no significant effect of average perceived
contact quality and no significant interaction term (for full results
see Table \ref{tab:intergroupGeneralTblWide}). We additionally used a
multilevel regression to check whether having an interaction with an
outgroup member had a situational (i.e., contemporaneous) effect within
the participants. We find that having an outgroup interaction is indeed
associated with significantly more positive outgroup attitudes within
the participants (random slopes model; \textit{b} = 2.99,
\textit{t}(4850) = 3.80, \textit{p} \textless{} .001,
\textit{95\%CI}{[}1.45, 4.54{]}), even after controlling for having an
interaction with a non-Dutch person (which did not relate to outgroup
attitudes independently; For full results see Table
\ref{tab:intergroupGeneralTblWide} and Online Supplementary Materials
B). Thus, in our second data set we find mixed results, outgroup
contacts show a positive effect on outgroup attitudes in the moment and
on average between participants. However, we find that this effect does
not depend on the average perceived interaction quality. This unexpected
results, might be due to the aggregation process and the following
analyses will focus in on the role of the perceived interaction quality
during an outgroup interaction.

@@

We again sequentially tested whether the fulfillment of the core need
during an interaction was (1) related to more positive outgroup
attitudes, (2) higher perceived interaction quality, and (3) whether the
variance explained by the core need is assumed by the perceived
interaction quality if considered jointly. We find that in the
multilevel models, the fulfillment of core situational needs during
outgroup contacts was associated with more positive outgroup attitudes
(random slopes model; \textit{b} = 0.13, \textit{t}(826) = 4.18,
\textit{p} \textless{} .001, \textit{95\%CI}{[}0.07, 0.19{]}) and also
predicted higher perceived interaction quality (random slopes model;
\textit{b} = 0.40, \textit{t}(826) = 6.46, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.28, 0.52{]}). Additionally, if we consider the
influences of core need fulfillment and interaction quality on outgroup
attitudes jointly, we find that virtually all variance is explained by
perceived interaction quality (random slopes mode; \textit{b} = 0.17,
\textit{t}(825) = 5.56, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.11, 0.23{]}) and only little unique variance is
still explained by core need fulfillment (\textit{b} = 0.03,
\textit{t}(825) = 1.47, \textit{p} = 0.141, \textit{95\%CI}{[}-0.01,
0.07{]}; also see Table \ref{tab:intergroupNeedsTblWide} for full
results). These results are consistent with the results in study one and
we thus find support for our hypotheses and can conclude that in this
data set the fulfillment of core situational needs had a significant
influence on outgroup attitudes and that this effect is likely explained
by its effect through perceived interaction quality.

@@

We again checked for alternative models. First, when considering
generalized situational core need fulfillment together with whether an
intergroup contact took place, we find that there is only a minuscule
main effect of core need fulfillment (random slopes model; \textit{b} =
-0.03, \textit{t}(4849) = -1.34, \textit{p} = 0.181,
\textit{95\%CI}{[}-0.08, 0.02{]}) but a stronger interaction effect of
core need fulfillment and outgroup contact (\textit{b} = 0.06,
\textit{t}(4849) = 3.03, \textit{p} = 0.002, \textit{95\%CI}{[}0.02,
0.10{]}). Together with a significant main effect of having an outgroup
contact (\textit{b} = 2.88, \textit{t}(4849) = 3.71, \textit{p}
\textless{} .001, \textit{95\%CI}{[}1.36, 4.40{]}), this indicates that
it is not key need fulfillment in general --- but key need fulfillment
during an outgroup contact that predicts more positive outgroup
attitudes. This finding is consistent with the results of the previous
study, albeit with slightly weaker effect (likely because of the large
number of measurements that did not include an outgroup interaction; For
full results see Table \ref{tab:intergroupNeedsTblWide} and for visual
comparison see Figure \ref{fig:AllportNeedFulfillment}).

In a final step we again checked whether during the interaction the core
situational need remains a meaningful predictor even when taking other
fundamental psychological needs into account. We find that the core need
adds significantly above a model with only the self determination theory
needs (random slopes models; \(\chi^2\)(6, \textit{N} = 108) = 22.90,
\textit{p} \textless{} .001). We find that the core need explains the
most variance in outgroup attitudes after an outgroup contact
(\textit{b} = 0.08, \textit{t}(823) = 2.95, \textit{p} = 0.003,
\textit{95\%CI}{[}0.03, 0.13{]}). When compared to the model with only
the SDT needs, the core need fulfillment flexibly takes on some of the
explained variance of all of the three fundamental needs. However,
different from the first study, relatedness (\textit{b} = 0.07,
\textit{t}(823) = 4.26, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.04, 0.10{]}) and autonomy (\textit{b} = 0.02,
\textit{t}(823) = 0.93, \textit{p} = 0.353, \textit{95\%CI}{[}-0.02,
0.08{]}) also predicted positive outgroup attitudes in this larger
sample. For full results see Table \ref{tab:intergroupNeedsTblWide} and
Online Supplementary Information B.

@@

As in Studies 1 and 2, we first conducted a general contact hypothesis
test. We found no correlation between intergroup contact and average
outgroup attitudes --- for neither the total number of outgroup
interactions nor the number of measurement beeps with an interaction
(\textit{r} \textless{} 0.13, \textit{p} \textgreater{} 0.291). However,
we did find a significant correlation between the participants' Average
Interaction Quality and their Average Outgroup Attitudes (\textit{r} =
0.25, \textit{p} = 0.036). When considering the number of interactions
and average interaction quality jointly in a linear regression, we only
found a main effect of average perceived interaction quality on outgroup
attitudes (\textit{b} = 0.36, \textit{t}(67) = 2.22, \textit{p} = 0.030,
\(\eta_p^2\) = 0.07) but no significant effect of the number of contacts
nor an interaction effect. This result mirrors that of study one but is
inconsistent with the second study. Additionally in a multilevel
regression to check for contemporaneous within person effects, we found
that having an outgroup interaction was indeed associated with
significantly more positive outgroup attitudes within the participants
(random slopes model; \textit{b} = 5.63, \textit{t}(3834) = 6.61,
\textit{p} \textless{} .001, \textit{95\%CI}{[}3.97, 7.29{]}), even
after controlling for having an interaction with a non-Dutch (which did
not relate to outgroup attitudes independently; for full results see
Table \ref{tab:intergroupGeneralTblWide} and Online Supplementary
Materials B). Thus, in our third data set we again found mixed results.
Outgroup contacts were on average not significantly related to outgroup
attitudes on a between participant level and this effect was also not
suppressed by an average interaction quality (i.e., no interaction
effect). However, while the aggregate data was partly inconsistent with
the first two studies, the within person contemporaneous effect of
intergroup contact was consistent across all three studies.

@@

We sequentially tested whether the fulfillment of Allport's Contact
Conditions was (1) related to more positive outgroup attitudes, (2)
higher perceived interaction quality, and (3) whether the variance
explained by Allport's Conditions is assumed by the perceived
interaction quality if considered jointly. We find that in the
multilevel models, the fulfillment of Allport's Conditions during
outgroup contacts was associated with more positive outgroup attitudes
(random slopes model; \textit{b} = 0.22, \textit{t}(1601) = 5.86,
\textit{p} \textless{} .001, \textit{95\%CI}{[}0.15, 0.29{]}) and also
predicted higher perceived interaction quality (random slopes model;
\textit{b} = 0.65, \textit{t}(1605) = 11.82, \textit{p} \textless{}
.001, \textit{95\%CI}{[}0.54, 0.76{]}). Moreover, when we consider the
influences of Allport's Conditions and interaction quality on outgroup
attitudes jointly, we find that perceived interaction quality is a
substantially stronger predictor (random slopes mode; \textit{b} = 0.17,
\textit{t}(1600) = 6.25, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.12, 0.23{]}) and the unique variance explained by
Allport's Conditions was less than half of its original effect size
(\textit{b} = 0.09, \textit{t}(1600) = 2.89, \textit{p} = 0.004,
\textit{95\%CI}{[}0.03, 0.15{]}; also see Table
\ref{tab:intergroupNeedsTblWide} and Figure
\ref{fig:AllportNeedFulfillment}). These results indicate that in this
dataset the the fulfillment of Allport's conditions had a significant
influence on outgroup attitudes and this effect is likely in parts
explained by its effect through perceived interaction quality.

@@

We again sequentially tested whether the fulfillment of the core need
during an interaction was (1) related to more positive outgroup
attitudes, (2) higher perceived interaction quality, and (3) whether the
variance explained by the core need is assumed by the perceived
interaction quality if considered jointly. We find that in the
multilevel models, the fulfillment of core need fulfillment during
outgroup contacts was associated with more positive outgroup attitudes
(random slopes model; \textit{b} = 0.19, \textit{t}(1601) = 5.29,
\textit{p} \textless{} .001, \textit{95\%CI}{[}0.12, 0.27{]}) and also
predicted higher perceived interaction quality (random slopes model;
\textit{b} = 0.43, \textit{t}(1605) = 8.35, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.33, 0.54{]}). Additionally, once we consider the
influences of core need fulfillment and interaction quality on outgroup
attitudes jointly, we find that perceived interaction quality is a
substantially stronger predictor (random slopes mode; \textit{b} = 0.17,
\textit{t}(1600) = 6.86, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.12, 0.22{]}) and the unique variance explained by
core need fulfillment was roughly half of its original effect size
(\textit{b} = 0.11, \textit{t}(1600) = 3.50, \textit{p} \textless{}
.001, \textit{95\%CI}{[}0.05, 0.17{]}; for full results see Table
\ref{tab:intergroupNeedsTblWide} and Figure
\ref{fig:AllportNeedFulfillment}). These results indicate that in this
data set the the fulfillment of core situational needs had a significant
influence on outgroup attitudes and this effect is likely in parts
explained by its effect through perceived interaction quality.

@@

To test wether Allport's conditions or the core need fulfillment are
better at predicting outgroup attitudes, we first assess relative model
performance indices (i.e., Akaike information criterion, and Bayesian
information criterion), and then consider the two predictors in a joint
model to see whether the two approaches predict the same variance in
outgroup attitudes. When comparing the model selection indices we find
that the fulfillment of the situational core need, indeed performs
slightly better than the model using Allport's conditions
(\(AIC_{CoreNeed}\) 12632.02 \textless{} 12651.59 \(AIC_{Allport}\), and
\(BIC_{CoreNeed}\) 12664.55 \textless{} 12684.12 \(BIC_{Allport}\)).
Additionally, when considering the predictors jointly, we find that both
significantly predict outgroup attitudes with similar sized regression
parameters (random slopes model; Allport's Conditions: \textit{b} =
0.16, \textit{t}(1600) = 4.92, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.09, 0.24{]}, Core Need: \textit{b} = 0.14,
\textit{t}(1600) = 3.85, \textit{p} \textless{} .001,
\textit{95\%CI}{[}0.08, 0.17{]}; also see Table
\ref{tab:intergroupNeedsTblWide} and Figure
\ref{fig:AllportNeedFulfillment}). This indicates that, although both
Allport's conditions and the core need fulfillment seem to (in part)
work through perceived interaction quality, they explain different
aspects of the variance in outgroup attitudes and do not constitute one
another.
